---
title: "QBS 181 Final Project Aim 2 (Say OK)"
author: "Say OK"
date: "2024-10-28"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Libraries
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(tidyr)
library(readxl)
library(ppcor)
library(lme4)
library(knitr)
```



# I. Load and Merge Datasets

## 1. Load the cleaned agricultural dataset (from Excel worksheet)
Please refer to the excel file for details on data cleaning, calculations, and exploratory steps.
We created a sheet named "R", which includes all columns needed for input into R.
```{r}
# Set working directory ("Session" --> "Set Working Directory")
# Input the sheet named "R" in the Excel file
ag_data <- read_excel("/Users/violet/Documents/QBS Fall 2024/QBS 181/Final Project/Data/Ag_Data_Clean.xlsx", sheet = "R")

head(ag_data)
dim(ag_data)
```


## 2. Load 2022 Parkinson mortality rate data (downloaded from CDC)
Longitudinal data on Parkinson's disease mortality rate over years can be downloaded directly from CDC.
All rates are age-adjusted mortality rates and represent the number of deaths per 100,000 total population.
```{r}
# Set working directory ("Session" --> "Set Working Directory")
# Input the sheet named "R" in the Excel file
mortality <- read.csv("/Users/violet/Documents/QBS Fall 2024/QBS 181/Final Project/Data/Parkinson_Mortality_2022.csv")
# Remove the last column (unnecessary URL)
mortality <- mortality[, -ncol(mortality)]

head(mortality)
dim(mortality)
```

Filter to keep 2022
```{r}
# Filter to keep year == 2022
mortality <- mortality %>%
  filter(YEAR == 2022)

dim(mortality)
```


## 3. Combine agricultural dataset with 2022 Parkinson mortality rate data
Create a conversion table as state is presented as full name in ag_data whereas it is presented as abbreviation in the mortality.
```{r}
# Create the conversion table as a data frame
state_abbreviations <- data.frame(
  State = toupper(c("Alabama", "Alaska", "Arizona", "Arkansas", "California", 
                    "Colorado", "Connecticut", "Delaware", "Florida", "Georgia",
                    "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", 
                    "Kentucky", "Louisiana", "Maine", "Maryland", 
                    "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
                    "Missouri", "Montana", "Nebraska", "Nevada", 
                    "New Hampshire", "New Jersey", "New Mexico", "New York", 
                    "North Carolina", "North Dakota", "Ohio", "Oklahoma", 
                    "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", 
                    "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", 
                    "Virginia", "Washington", "West Virginia", "Wisconsin", 
                    "Wyoming")),
  State_Abbrev = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                  "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                  "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                  "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                  "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY")
)
head(state_abbreviations)
```

Merge the datasets
```{r}
ag_data <- merge(ag_data, state_abbreviations, by.x = "State_Name", by.y = "State", all.x = TRUE)
df <- merge(ag_data, mortality, by.x = "State_Abbrev", by.y = "STATE", all.x = TRUE)

head(df)
```



# II. Aim 2a: Investigate the relationship between PD mortality rates and total agricultural chemical expenditure per state.

## 1. PD mortality rates vs. Total agricultural chemical expenditure per state
a. scatter plot visualization
```{r}
ggplot(df, aes(x = Total_Chemical_Expenditure, y = RATE)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = TRUE) + 
  labs(title = "Scatter Plot of PD Mortality Rates vs. Total Chemical Expenditure",
       x = "Total Agricultural Chemical Expenditure",
       y = "PD Mortality Rate (per 100,000 population)") +
  theme_minimal()
```

b. Spearman's rank correlation for statistical testing
```{r}
spearman_correlation <- cor(df$Total_Chemical_Expenditure, df$RATE, method = "spearman", use = "complete.obs")
spearman_correlation
```
Result: There is a weak positive monotonic relationship between Total Chemical Expenditure and Mortality RATE.


## 2.PD mortality rates vs. Chemical Expenditure per Farmland Acres per State
a. scatter plot visualization
```{r}
ggplot(df, aes(x = Chemical_Expenditure_per_Acres, y = RATE)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = TRUE) + 
  labs(title = "Scatter Plot of PD Mortality Rates vs. Chemical Expenditure per Farmland Acres",
       x = "Chemical Expenditure per Farmland Acres",
       y = "PD Mortality Rate (per 100,000 population)") +
  theme_minimal()
```

b. Spearman's rank correlation for statistical testing
```{r}
spearman_correlation <- cor(df$Chemical_Expenditure_per_Acres, df$RATE, 
                            method = "spearman", use = "complete.obs")
spearman_correlation
```
Result: There is a weak inverse relationship, meaning that as agricultural chemical expenditure rises, Parkinson’s mortality rate slightly decreases.


## 3. PD mortality rates vs. Total Chemical Expenditure Considering Geographic Regions
a. boxplot and heatmap
```{r}
# Box Plot: PD mortality rates by Geographic Region
ggplot(df, aes(x = Geographic_Region, y = RATE)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  labs(title = "Box Plot of PD Mortality Rates by Geographic Region",
       x = "Geographic Region",
       y = "PD Mortality Rate (per 100,000 population)") +
  theme_minimal()

# Heatmap: PD mortality rate by Geographic Divisions and Total Chemical Expenditures
## Divide the Chemical_Expenditure_per_Acres into quintile
df <- df %>%
  mutate(Expenditure_Bin = cut(Total_Chemical_Expenditure, breaks = 5))
## Plot
ggplot(df, aes(x = Geographic_Division, y = Expenditure_Bin, fill = RATE)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Heatmap of PD Mortality Rates by Region and Expenditure Levels",
       x = "Geographic Region",
       y = "Total Chemical Expenditure (Binned)",
       fill = "PD Mortality Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

b. Fit the mixed-effects model
Since the Geographic Region is a higher-level above State, we can use a mixed-effects model to show a more accurate relationship
```{r}
# Fit the mixed-effects model
# Assuming RATE is the dependent variable 
# and Total_Chemical_Expenditure is the main predictor
model <- lmer(RATE ~ Total_Chemical_Expenditure + (1 | Geographic_Region)
              , data = df)

# Summary of the model
summary(model)

# Check normality of residuals of the dependent variable
## Fit a simple linear regression model
model <- lm(RATE ~ Total_Chemical_Expenditure, data = df)
summary(model)
## Extract residuals
residuals <- resid(model)
qqnorm(residuals)
qqline(residuals, col = "red")
```
Interpretation: The coefficient indicates that for each dollar increase in Total Chemical Expenditure, the RATE (PD mortality rate) is expected to increase by 3.709e-10 per 100,000 person. In other word, for each billion-dollar increase in Total Chemical Expenditure, the RATE (PD mortality rate) is expected to increase by 0.3709 deaths per 100,000 people. The p-value for Total_Chemical_Expenditure is 0.241, which is above the significance level (0.05). This means the effect of Total Chemical Expenditure on PD mortality rate is not statistically significant in this model.



# III. Aim 2b: Evaluate the association between the amount of acreage treated with three types of pesticides and PD mortality rates.

### Look at totals and averages
```{r}
# create a summary table of the total acreage of treated farmland 
# for different types of pesticides and the mortality rate
summary_table <- df %>%
  summarise(
    Insecticide_No_Nema_Sum_Acres = 
      sum(Insecticide_No_Nema_Treatment_Acres, na.rm = TRUE),
    Insecticide_No_Nema_Avg_Percentage = 
      mean(`Insecticide_No_Nema_Treatment_Percentage_(%)`, na.rm = TRUE),
    Insecticide_No_Nema_SD_Percentage = 
      sd(`Insecticide_No_Nema_Treatment_Percentage_(%)`, na.rm = TRUE),
    
    Insecticide_Nema_Sum_Acres = 
      sum(Insecticide_Nema_Treatment_Acres, na.rm = TRUE),
    Insecticide_Nema_Avg_Percentage = 
      mean(`Insecticide_Nema_Treatment_Percentage_(%)`, na.rm = TRUE),
    Insecticide_Nema_SD_Percentage = 
      sd(`Insecticide_Nema_Treatment_Percentage_(%)`, na.rm = TRUE),
    
    Herbicide_Sum_Acres = 
      sum(Herbicide_Treatment_Acres, na.rm = TRUE),
    Herbicide_Avg_Percentage = 
      mean(`Herbicide_Treatment_Percentage_(%)`, na.rm = TRUE),
    Herbicide_SD_Percentage = 
      sd(`Herbicide_Treatment_Percentage_(%)`, na.rm = TRUE),
    
    Fungicide_Sum_Acres = 
      sum(Fungicide_Treatment_Acres, na.rm = TRUE),
    Fungicide_Avg_Percentage = 
      mean(`Fungicide_Treatment_Percentage_(%)`, na.rm = TRUE),
    Fungicide_SD_Percentage = 
      sd(`Fungicide_Treatment_Percentage_(%)`, na.rm = TRUE),
    
    Mortality_Sum_Rate = sum(RATE, na.rm = TRUE),
    Mortality_Avg_Rate = mean(RATE, na.rm = TRUE),
    Mortality_SD_Rate = sd(RATE, na.rm = TRUE)
  ) 

print(summary_table)

(correlations <- data.frame(
  Treatment_Type = c("Insecticide No Nema Acres", 
                     "Insecticide No Nema Percentage",
                     "Insecticide Nema Acres", 
                     "Insecticide Nema Percentage",
                     "Herbicide Acres", 
                     "Herbicide Percentage",
                     "Fungicide Acres", 
                     "Fungicide Percentage"),
  Spearman_Correlation_with_Mortality = c(
    cor(df$Insecticide_No_Nema_Treatment_Acres, df$RATE, 
        method = "spearman", use = "complete.obs"),
    cor(df$`Insecticide_No_Nema_Treatment_Percentage_(%)`, 
        df$RATE, method = "spearman", use = "complete.obs"),

    cor(df$Insecticide_Nema_Treatment_Acres, df$RATE, 
        method = "spearman", use = "complete.obs"),
    cor(df$`Insecticide_Nema_Treatment_Percentage_(%)`, 
        df$RATE, method = "spearman", use = "complete.obs"),

    cor(df$Herbicide_Treatment_Acres, df$RATE, 
        method = "spearman", use = "complete.obs"),
    cor(df$`Herbicide_Treatment_Percentage_(%)`, df$RATE, 
        method = "spearman", use = "complete.obs"),

    cor(df$Fungicide_Treatment_Acres, df$RATE, 
        method = "spearman", use = "complete.obs"),
    cor(df$`Fungicide_Treatment_Percentage_(%)`, 
        df$RATE, method = "spearman", use = "complete.obs")
  )
))

kable(correlations, caption = "Spearman Correlation Matrix")
```

```{r}
correlations_long <- correlations %>%
  pivot_longer(cols = Spearman_Correlation_with_Mortality, 
               names_to = "Correlation_Type", values_to = "Correlation_Value")

ggplot(correlations_long, aes(x = Treatment_Type, y = Correlation_Value, 
                              fill = Correlation_Value)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, name = "Spearman Correlation") +
  labs(
    title = "Spearman Correlation between Treatment and Mortality Rate",
    x = "Pesticide Treatment",
    y = "Spearman Correlation with Mortality Rate"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

Interpretation: There is no clear association between these pesticide treatments (measured as total acreage and percentage of farmland treated) with PD mortality rate.

## 

```{r}
cor_data <- df %>%
  dplyr::select(
    Total_Acres,
    Insecticide_No_Nema_Treatment_Acres,
    `Insecticide_No_Nema_Treatment_Percentage_(%)`,
    Insecticide_Nema_Treatment_Acres,
    `Insecticide_Nema_Treatment_Percentage_(%)`,
    Herbicide_Treatment_Acres,
    `Herbicide_Treatment_Percentage_(%)`,
    Fungicide_Treatment_Acres,
    `Fungicide_Treatment_Percentage_(%)`,
    DEATHS
  )

cor_data <- cor_data %>% mutate(across(everything(), as.numeric))

# Compute Spearman correlation matrix
cor_matrix <- cor(cor_data, method = "spearman", use = "complete.obs")

# Display the correlation matrix
print(cor_matrix)

colnames(cor_matrix) <- c(
  "Total Acres", "Insect No Nema Acres", "Insect No Nema %",
  "Insect Nema Acres", "Insect Nema %", "Herb Acres", "Herb %",
  "Fung Acres", "Fung %", "Deaths"
)
rownames(cor_matrix) <- c(
  "Total Acres", "Insect No Nema Acres", "Insect No Nema %",
  "Insect Nema Acres", "Insect Nema %", "Herb Acres", "Herb %",
  "Fung Acres", "Fung %", "Deaths"
)
kable(cor_matrix, caption = "Spearman Correlation Matrix")
```
very weak to weak correlation of some treatment measurements and PD mortality

